---
topic: sustainability
output: envreportutils.internal::print_ver
params:
  input_source:
    label: "Input Source:"
    value: local
    choices: [bcdc,local]
    input: select
---
<!--
Copyright 2025 Province of British Columbia

Licensed under the Apache License, Version 2.0 (the "License");
you may not use this file except in compliance with the License.
You may obtain a copy of the License at

http://www.apache.org/licenses/LICENSE-2.0

Unless required by applicable law or agreed to in writing, software distributed under the License is distributed on an "AS IS" BASIS,
WITHOUT WARRANTIES OR CONDITIONS OF ANY KIND, either express or implied.
See the License for the specific language governing permissions and limitations under the License.
-->


```{r set-options, echo=FALSE, message=FALSE, cache=FALSE, warning=FALSE}
# Set the 'Input Source' parameter value (either directly in the header
# of this file, or through the 'Knit with parameters' menu in RStudio.
# Set it to to 'bcdc' if you want to build using data sourced from the BC Data 
# Catalogue, or to 'local' if you want to build it directly from the output of 
# running 01_load.R and 02_output.R

library(dplyr)
library(readr)
library(here)

input <- if (params$input_source == "bcdc") {
  "https://catalogue.data.gov.bc.ca/dataset/d21ed158-0ac7-4afd-a03b-ce22df0096bc/resource/d2648733-e484-40f2-b589-48192c16686b/download/bc_municipal_solid_waste_disposal.csv"
} else if (params$input_source == "local") {
  here("out", "BC_Municipal_Solid_Waste_Disposal.csv")
} else {
  stop("Invalid input source")
}

msw <- read_csv(input)

myIdent <- function(x) {
  x <- ifelse(is.na(x),"N/A",as.character(x))
}

updateDate <- format(Sys.Date(), "%B %Y")

max_year <- max(msw$Year, na.rm = TRUE)

BC_data <- msw %>% 
  group_by(Year) %>% 
  summarise(Total_Disposed_Tonnes = sum(Total_Disposed_Tonnes, na.rm = TRUE), 
            Population = sum(Population, na.rm = TRUE), 
            Disposal_Rate_kg = round(Total_Disposed_Tonnes / Population * 1000)) %>%
  mutate(Total_Disposed_Tonnes = case_when (
    Year == 2023 ~ (Total_Disposed_Tonnes + 50000), # Adjust 2023 data by 50k. Note: there is a 50K tonnes of unclaimed waste (FV claimed it’s from MV) that has been added to the total but it’s not reflected in any RD’s tonnage.
    TRUE ~ Total_Disposed_Tonnes),
         Disposal_Rate_kg = case_when (
           Year == 2023 ~ round((Total_Disposed_Tonnes) / Population * 1000), # Recalculate 2023 rate after adjustment
           TRUE ~ Disposal_Rate_kg))

max_year_rate <- BC_data$Disposal_Rate_kg[BC_data$Year == max_year]

max_year_pop <- BC_data$Population[BC_data$Year == max_year]

max_year_total <- BC_data$Total_Disposed_Tonnes[BC_data$Year == max_year]

comp_year <- 2012

comp_year_rate <- BC_data$Disposal_Rate_kg[BC_data$Year == comp_year]

min_rate <- min(BC_data$Disposal_Rate_kg, na.rm = TRUE)

min_rate_year <- max(BC_data$Year[BC_data$Disposal_Rate_kg == min_rate], na.rm = TRUE)
```

---
title: "`r sprintf('Municipal Solid Waste Disposal in B.C. (1990-%s)', max_year)`"
---

This indicator reports the estimated amount of municipal solid waste
disposed per person for each regional district and for British Columbia.
The per person disposal rate is an estimate of how many kilograms of solid
waste each B.C. resident sends to a landfill or other disposal site in a
given year.

-   Approximately `r round(max_year_total / 1e6, 1)` million tonnes of municipal 
    solid waste was disposed of in B.C. in `r max_year`. While there has been an overall
    decreasing trend in municipal solid waste disposal rates since the
    1990s, much of the waste disposed of could be reduced (by not
    being generated in the first place), repurposed, composted or
    recycled.
-   Waste reduction efforts are more and more important as [B.C. population continues to
    increase](http://www.env.gov.bc.ca/soe/indicators/sustainability/bc-population.html),
    landfills approach their capacity for receiving new waste, and the
    imperative to reduce greenhouse gas emissions from waste disposal
    grows.
-   Tracking waste disposal across B.C. is important as it provides
    historical trend information and the current status of waste
    disposal---both of which inform and guide efforts to reduce the
    amount of waste we create.
-   In `r max_year`, British Columbians disposed of an
    average of `r round(max_year_rate)` kg of municipal solid
    waste per person---down `r abs(max_year_rate - comp_year_rate)` kg per person since `r comp_year`, though higher 
    than the low of `r min_rate` kg/person in `r min_rate_year`.
-   Provincial initiatives which support decreases in waste disposal
    rates include the [CleanBC Plastics Action Plan](https://www2.gov.bc.ca/gov/content?id=4729A52504704E94A74FC690C907D7F9), [organics
    diversion](https://www2.gov.bc.ca/gov/content/environment/waste-management/food-and-organic-waste/organic-waste-diversion), [food
    waste
    prevention](https://www2.gov.bc.ca/gov/content/environment/waste-management/food-and-organic-waste/prevent-food-waste),
    [extended producer
    responsibility (recycling) programs](https://www2.gov.bc.ca/gov/content?id=BFAE6EF8C6E6443A9B2131D447E456DF), and [funding for organics infrastructure and collection](https://www2.gov.bc.ca/gov/content?id=715397A83DFC4A75BC21B810012A5F52).
-   Variability in disposal rates from one region to the next is
    influenced by factors such as population density, economic activity,
    tourist and transient population fluctuations, distance to recycling
    markets, the role of various stewardship agencies operating in an
    area and the capacity and infrastructure in place in a region. While disposal rate data do not tell the whole story, they are useful in
    setting the stage for continuous improvement in waste management in
    B.C.

The charts below show the estimated per person weight of solid waste
disposed of in `r max_year` for each regional district and the province as a
whole, as well as a history of disposal rates for each region.

```{r bc-plot, fig.height=3.5, echo=FALSE, message=FALSE, warning=FALSE, error=TRUE}
library(ggplot2)
library(grid)

p <- ggplot(data = BC_data, aes(x = Year, y = Disposal_Rate_kg))
bars <- geom_bar(stat = 'identity', fill = '#08519c',
                 width = 0.6)
labs <- labs(title = paste0("Municipal Solid Waste disposal rates in British Columbia, 1990-", max_year)
             , y = "Disposal Rate (kg / person)")
theme <- theme(panel.background = element_blank()
               , panel.grid.major.y = element_line(colour = "darkgray", size = 0.5)
               , panel.grid.minor.y = element_line(colour = "lightgray", size = 0.5)
               , panel.grid.major.x = element_blank()
               , panel.grid.minor.x = element_blank()
               , axis.ticks.y = element_blank()
               , plot.margin = unit(c(1.5,2,0.5,1), "lines")
               , plot.title = element_text(size = 12)
               , axis.text.x = element_text(size = 9, angle = 0)
               , axis.text.y = element_text(size = 9)
               , axis.title.x = element_text(vjust = 0, size = 11)
               , axis.title.y = element_text(hjust = 0.5
                                             , vjust = 1
                                             , angle = 90
                                             , size = 11))

scale.y <- scale_y_continuous(breaks = seq(0, max(BC_data$Disposal_Rate_kg
                                                  , na.rm = TRUE) + 100, by = 100)
                              , limits = c(0, max(BC_data$Disposal_Rate_kg
                                                  , na.rm = TRUE) + 50)
                              , expand = c(0,0))

#scale.x <- scale_x_continuous(breaks = c(seq(min(BC_data$Year), max(BC_data$Year)-3, by = 5), 2023))
#scale.x <- scale_x_continuous(breaks = sort(c(seq(max(BC_data$Year), min(BC_data$Year)+8, by = -5), 1994, 1990)))
scale.x <- scale_x_continuous(breaks = c(seq(min(BC_data$Year), max(BC_data$Year)-5, by = 4), 2023))

p + bars + labs + theme + scale.y + scale.x
```

```{r rd-plots, echo=FALSE, message=FALSE, warning=FALSE, error=TRUE, fig.width = 8, fig.height=11}
library(RColorBrewer)

ordered_by_pop <- msw$Regional_District[msw$Year == max_year][order(msw$Population[msw$Year == max_year])]

msw$Regional_District <- factor(msw$Regional_District, levels = as.character(ordered_by_pop))

msw$Population_Category <- cut(msw$Population, 
                               breaks = c(0, 10000, 50000, 100000, 500000, 1e6, Inf), 
                               labels = c("< 10,000", "< 50,000", "< 100,000", "< 500,000", 
                                          "< 1,000,000", "> 1,000,000"), 
                               include.lowest = TRUE)

ggplot(msw, aes(x = Year, y = Disposal_Rate_kg, fill = Population_Category)) +
  geom_bar(stat = "identity", width = 0.6) +
  scale_fill_manual(values = brewer.pal(9, "Blues")[5:9]) + 
  facet_wrap(~Regional_District, ncol = 4, scales = "free_x") + 
  labs(title = paste0("Municipal Solid Waste disposal rates in B.C. Regional Districts, 1990-", max_year, "\n"), 
       y = "Disposal Rate (kg / person)", fill = "Regional District\nPopulation") + 
  theme + 
  theme(legend.position = "bottom", legend.key.width = unit(1, "cm"), 
        strip.text = element_text(size = 8),
        axis.text = element_text(size = 4.5), 
        axis.text.x = element_text(angle = 45, hjust = 0.8), 
        plot.margin = unit(c(0,0,0,0), "mm")) + 
  scale_x_continuous(breaks = c(seq(1990, 2018, by = 4), 2023))
```

\begin{center}
\fbox{
\begin{minipage}[c]{12cm}
\textbf{What is Included in Disposal Rate?}
\vspace{-3.5mm}
\begin{itemize}
\itemsep1pt\parskip0pt\parsep0pt
\item
Disposal rates reported here include waste from the residential sector, 
institutional, commercial, and light industrial sources as well as 
waste from construction, renovation, and demolition activities.
\item
Disposal rates do not include hazardous, biomedical, or agricultural waste, 
motor vehicles or components, heavy industry, or contaminated soil.
\item
Waste that is reused or recycled is also not included.
\end{itemize}
\end{minipage}
}
\end{center}


#### Interpreting the Data:

-   `r max_year` disposal rates were provided by regional districts.
-   Where no bar is shown on a chart for certain previous years, a
    verifiable estimate was not available.
-   Historic disposal rates cannot be directly compared between years
    because the data were collected using different methods. However,
    the data from `r comp_year` onward have been collected using the same
    methodology.
-   Regional district population estimates are sourced from [BC
    Stats](https://www2.gov.bc.ca/gov/content?id=36D1A7A4BEE248598281824C13CB65B6).
    Alberni-Clayoquot, Metro Vancouver, Squamish-Lillooet, and East 
    Kootenay Regional Districts and Northern Rockies Regional Municipality population estimates have been 
    adjusted to take into account tourist populations, 
    based on data submitted by the regional districts.
-   Disposal rates for Comox Valley and Strathcona are reported as one
    as they manage their waste
    streams [together](http://www.cswm.ca/).

Regional districts that wish to update the waste disposal data reported
above should contact the Environmental Policy and Initiatives Branch. Input from regional districts will
also serve to inform future waste reporting methodology development. 

The  [**R**](http://www.r-project.org) code supporting this indicator is [available on GitHub](https://github.com/bcgov/msw-disposal-indicator).

### References and Other Useful Links:

-   Find out where to recycle using the Recycling Council of British
    Columbia's [Recyclepedia
    App](https://www.rcbc.ca/recyclepedia/search)

-   [B.C.'s recycling system (EPR) and regulations
    ](https://www2.gov.bc.ca/gov/content/environment/waste-management/recycling/recycling-system-regulations)

-   [Solid Waste Management Planning in B.C.                                                                          ](http://www2.gov.bc.ca/gov/content?id=CBCECCAABAA2401B9FB3B7CE63129709)

-   [British Columbia\'s *Environmental Management
    Act*](http://www.bclaws.ca/civix/document/id/complete/statreg/03053_00)

-   [British Columbia Ministry of Environment and Climate Change Strategy 2025/26 – 2027/28 Service Plan (March 2025) (PDF)](https://www.bcbudget.gov.bc.ca/2025/sp/pdf/ministry/env.pdf)

### Data
  
\*By accessing these datasets, you agree to the licence associated with
each file, as indicated in parentheses below.

-   [Municipal Solid Waste Disposal,
1990-`r max_year`](https://catalogue.data.gov.bc.ca/dataset/d21ed158-0ac7-4afd-a03b-ce22df0096bc) (Licence: [Open Government Licence - British Columbia](http://www2.gov.bc.ca/gov/content?id=A519A56BC2BF44E4A008B33FCF527F61))

Published and Available On-Line at Environmental Reporting BC (`r format(Sys.Date()+10, "%B %Y")`):
<http://www.env.gov.bc.ca/soe/indicators/sustainability/municipal-solid-waste.html>